---
title: "Tutorial of Mixture DEG Analysis (SpotGLM) on Barrett's Esophagus example"
author: "Jiazhen Rong, Sydney Bracht"
date: "2026-06-09"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SUMMIT Mixture DEG Tutorial}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = "~/Documents/GitHub/SpatialMT/examples/")

library(SUMMIT)
library(SummarizedExperiment)
library(Matrix)
library(dplyr)
library(ggplot2)
library(Seurat)
library(spacexr)
```

## Overview

In the [main SUMMIT tutorial](Tutorial_a200_s6.html), we tested **which
specific cell type carries a given mitochondrial variant**. Once we know
which spots are carriers of a clone, a natural follow-up question is:

> *Which genes are differentially expressed between clones (or between a
> clone and the rest of the tissue, or a clone and its microenvironment)?*

In this tutorial, our goal is to identify genes that are differentially expressed in a cell type–specific manner, either within a clone or in its surrounding spatial neighborhood. For example, when a variant has been identified as tumor-associated, two key questions often arise: (1) How do tumor cells carrying the variant differ from other tumor cells? and (2) How does the immune and stromal microenvironment surrounding variant-carrying tumor cells differ from that surrounding tumor cells without the variant?

A central challenge in addressing these questions is that spatial transcriptomic spots typically contain mixtures of cell types. A naïve differential expression analysis (like standard single-cell DEG analysis) comparing spots with and without the variant would therefore conflate cell type–specific effects with differences in cellular composition. To account for this, we propose a mixture Poisson generalized linear model (GLM) that explicitly models spot-level cell type mixing in differential expression analysis.

For gene g and spot s, we assume the observed gene expression counts Y_sg follows a Poisson GLM model: 
$$Y_{sg} \sim Poisson(\lambda_{sg})$$
$$\lambda_{sg}=  \sum_k W_{sk} exp (X_s^T\beta_{gk}+L_s). $$     
where $L_s$ is the library size of spot $s$, $W_{sk}$ is the mixture proportion of cell type $k$ in spot $s$, which can be pre-computed from tools like RCTD through deconvolution.  
$X$ is the pre-designed covariate matrix, representing the identity of the clones. For example, a simple design for spot s and clone c could be:

$$X_{s,c}=\begin{cases} 1, & \text{if spot s belongs to clone c} \\ 0, & \text{if spot s does not belong to clone c} \end{cases}$$
From the model, we can estimate gene $g$ and celltype $k$ specific effect sizes $\beta_gk$. For further derivation and discussion of how to estimate the effect sizes $\beta_gk$, and how to test the effect sizes, please check our Methods section of **Mixture GLM Model for Differentially Expressions of Variants** in the manuscript.

This tutorial reuses the same Barrett's Esophagus sample (A200_s6) from
the main tutorial, takes the mitochondrial clones we identified there as
the covariate of interest (niche), and gives an example of the
mixture-DEG analysis.

The example data are available at the same
[DropBox](https://www.dropbox.com/scl/fo/sx6tfijkkapsvo4036xhj/AB9NjFsN6yz5xYyz-CpAgKk?rlkey=kf6blkddohhd7gdr96q85kf21&st=sybpv5mx&dl=0)
folder; please move them to the *SpatialMT/examples/example_data*
folder.

To run this tutorial you additionally need the
[spotglm](https://kaishumason.github.io/SpotGLM/) package:

```{r install_spotglm, eval=FALSE}
install.packages("devtools")
devtools::install_github("kaishumason/SpotGLM")
devtools::install_github("kaishumason/SPARROW") 
```

### Step 1: Load example data

We load the same three inputs as in the main tutorial:

-   the Spatial Transcriptomics Seurat object
-   the mtDNA variant allele frequency matrix from SUMMIT
-   the RCTD cell-type deconvolution matrix

```{r STEP1_lib, eval=T}
library(SUMMIT)               # SUMMIT
library(spotglm)                 # mixture GLM
library(spacexr)                 # RCTD
library(Seurat)
library(Matrix)
library(dplyr)
library(purrr)
library(tidyr)
library(tibble)
library(ggplot2)
library(grid)
library(gridExtra)
library(pheatmap)

setwd("~/Documents/GitHub/SpatialMT/examples/")
source("utility_prev_literature/210215_FunctionsGeneral.R") # from MAESTER paper
```

```{r STEP1_load_ST_data, eval=T}
slide_name = "A200_s6"

# load Spatial Transcriptomics / Curio Seeker data
seu <- readRDS("example_data/a200_s6_final.rds")
# spatial coordinates
coords = seu@reductions$SPATIAL@cell.embeddings
```

Load the RCTD cell-type ratio and normalize the per-spot weights to sum
to 1:

```{r STEP1_load_celltype_ratio, eval=T}
rctd_ratio_major = readRDS("example_data/RCTD_res.rds")
rctd_ratio_major = as.data.frame(as.matrix(normalize_weights(rctd_ratio_major@results$weights)))
rownames(rctd_ratio_major) = paste0(rownames(rctd_ratio_major), "-1")
deconv = rctd_ratio_major
```

Load the mtDNA data (variant allele frequency `af.dm` and coverage `N`):

```{r STEP1_load_MT_data, eval=T}
maegatk.rse = readRDS("example_data/maegatk_mr1.rds")
# all possible mitochondrial mutations' (4 * 16K) x spot VAF
af.dm <- data.matrix(computeAFMutMatrix(maegatk.rse)) # 0-1 VAF scale
# coverage N, spot x each chrM location's coverage
counts = as.matrix(maegatk.rse@assays@data$coverage)
rownames(counts) = 1:nrow(counts)
colnames(counts) = maegatk.rse@colData@rownames
N = as(as.matrix(counts[sapply(strsplit(rownames(af.dm), "_"), "[[", 1), ]), "sparseMatrix")
rownames(N) <- rownames(af.dm)
```

### Step 2: Build the clone covariate (niche) matrix

The "niche" matrix is the design/covariate matrix `X` that represents
different clone identity and their microenvironment. In this tutorial
example, each column is a mitochondrial clone, and a spot is assigned to
a clone if its VAF for that variant exceeds a heteroplasmy threshold
(0.25). Spots not carrying any of the variants of interest are assigned
to a shared `Rest` group, which serves as the reference background.

The covariate/niche matrix could be further designed to incorporate
additional columns such as the ratio of different micro-environment celltypes, etc. 

```{r STEP2_niche, eval=T}
# variants of interest (the same clones from the main tutorial)
voi = c("3054_G>C", "15777_G>C", "3071_T>C")

niche = as.matrix(Matrix(0, nrow = dim(seu)[2], ncol = 4)) # X covariate matrix
rownames(niche) = colnames(seu)
colnames(niche) = c("3054_G>C", "15777_G>C", "3071_T>C", "Rest")

# assign carrier spots (VAF > 0.25) for each clone
niche[colnames(af.dm)[af.dm["3054_G>C", ] > 0.25], "3054_G>C"] = 1
niche[intersect(colnames(af.dm)[af.dm["15777_G>C", ] > 0.25], rownames(niche)), "15777_G>C"] = 1
niche[intersect(colnames(af.dm)[af.dm["3071_T>C", ] > 0.25], rownames(niche)), "3071_T>C"] = 1
# everything else is background ("Rest")
niche[(niche[, "3054_G>C"] == 0) & (niche[, "15777_G>C"] == 0) & (niche[, "3071_T>C"] == 0),
      "Rest"] = 1

colSums(niche) # number of spots per clone
```

Prepare the input data as the following format:

-   `counts` : spot x gene raw UMI counts (such as the top 2000 highly
    variable genes),
-   `deconv` : spot x cell-type proportion matrix $W$,
-   `niche` : spot x clone covariate matrix $X$ (built as above),
-   `library_size`: per-spot total counts (used as the GLM offset),
-   `coords` : spatial coordinates of each spot.

```{r STEP3_assemble, eval=T}
data <- list()
inter_bc = intersect(colnames(seu), rownames(deconv))

data$coords = coords[inter_bc, ]
data$deconv = deconv[inter_bc, ]

# Subset to the 2000 most variable genes
seu <- FindVariableFeatures(seu, selection.method = "vst", nfeatures = 2000)
data$counts = t(seu@assays$RNA@counts)[inter_bc, VariableFeatures(seu)]
data$library_size = rowSums(data$counts)[inter_bc]
data$niche = as.matrix(niche[inter_bc, ])

rm(af.dm); rm(maegatk.rse); gc()
```

### Step 3: Fit the mixture GLM

We fit a **spot-level Poisson GLM** for each gene. The model regresses
the gene's count in each spot on the clone covariate `X`, scaled by the
cell-type proportions `lambda = W`, with the log library size as an
offset. This deconvolves the mixed signal so that each clone's effect is
estimated in a **cell-type-specific** manner.

> *Fitting 2000 genes takes a while (\~ 1 hour on a standard laptop's
> cpu). For this tutorial we will directly load the pre-computed result
> spotglm_res_a200_s6.rds.*

```{r STEP4_patch_grad, eval=T}
# modification of fitting package
g_new <- function(X, y, beta, lambda, offset) {
  means <- spotglm:::spot_poisson[["predict"]](X, beta, lambda, offset)
  CT_means <- means$individual
  total_means <- means$total
  weights <- -lambda * CT_means + y * (1 / total_means) * lambda * CT_means
  grad_beta <- t(as.matrix(X)) %*% as.matrix(weights)
  list(grad = grad_beta, weights = weights)
}
assignInNamespace("spot_poisson",
                  within(spotglm:::spot_poisson, { grad <- g_new }),
                  ns = "spotglm")
```

```{r STEP4_load_fit, eval=F}
# load the pre-computed spotGLM fit (list of length 2000, one element per gene)
res <- readRDS("example_data/spotglm_res_a200_s6.rds")
length(res)
names(res[[1]]) # beta_estimate, vcov, likelihood, converged, ...
```

The following code was used to generate *spotglm_res_a200_s6.rds* object. Expected run time is ~ 1 hour. Can be skipped in this tutorial for time saving purpose. 

```{r STEP4_fit, eval=FALSE}
# Fitting for the 2000 HVGs. Expect ~ 1 hr to finish
num_genes = ncol(data$counts)
res = vector("list", num_genes)
t1 = Sys.time()
for (j in seq_len(num_genes)) {
  if (j %% 100 == 0) {
    cat("Fitting model for gene ", j, " out of ", num_genes, "\n")
    print(Sys.time() - t1)
  }
  res[[j]] = spotglm::run_model(y = data$counts[, j],
                                X = data$niche,
                                lambda = data$deconv,
                                family = "spot poisson",
                                offset = log(data$library_size),
                                initialization = TRUE, batch_size = 250)
}
names(res) = colnames(data$counts)
saveRDS(res, "example_data/spotglm_res_a200_s6.rds")
```


### Step 4: Mixture GLM based DEG

Each gene's fit gives a clone-specific coefficient (`beta_estimate`) and
its covariance (`vcov`) for every cell type. To perform weighted DEG
analysis regarding each celltype, we test the contrasts between clones, i.e. whether a gene's
expression differs between two clones *within a given cell type*. We use
`compute_contrast_significance()` and loop over all cell types $\times$
all clone pairs.

We run a **two-sided** test (`sided = 2`) so genes up- or down-regulated
in either clone are detected.

# You could also directly load the fitted result to save time:
```{r STEP5_contrasts_load_precomputed_result, eval=T}
cell_types <- colnames(deconv) # e.g. BE, SQ, IM, FB, VC, ...
niches <- colnames(niche)      # the 3 clone + Rest covariates
pairs <- combn(niches, 2, simplify = FALSE)  # all clone-vs-clone contrasts

all_contrasts <- readRDS("example_data/results/deg_2sided.rds")
```

The following code was used to generate *deg_2sided.rds* object. Can be skipped in this tutorial for time saving purpose. 
```{r STEP5_contrasts, eval=F}

all_contrasts <- purrr::map_dfr(cell_types, function(ct) {
  purrr::map_dfr(pairs, function(p) {
    n1 <- p[1]; n2 <- p[2]
    out <- compute_contrast_significance(
      input_list = res,
      cell_type = ct,
      effect_names = c(n1, n2),
      beta_name = "beta_estimate",
      covariance_name = "vcov",
      sided = 2
    )
    if (!("gene" %in% names(out))) out <- out %>% tibble::rownames_to_column("gene")
    out %>% mutate(contrast = paste0(n1, " vs ", n2))
  })
})

saveRDS(all_contrasts, "example_data/results/deg_2sided.rds")

head(all_contrasts)
```

The resulting table has one row per `gene` x `cell_type` x `contrast`,
with a `test_statistic`, raw `pval` and FDR-adjusted `qval`.

We typically drop mitochondrial (`MT-`) genes from the DEG results,
since they directly track the mtDNA clone definition and would otherwise
dominate the list:

```{r STEP5_filter_mt, eval=F}
all_contrasts = all_contrasts[!grepl("MT-", all_contrasts$name), ]
```

Quick summary of how many significant genes (q \< 0.05 / 0.10) each cell
type x contrast has:

```{r STEP5_summary, eval=F}
summary_by_ct_contrast <- all_contrasts %>%
  group_by(cell_type, contrast) %>%
  summarise(
    n_q05 = n_distinct(gene[qval < 0.05]),
    n_q10 = n_distinct(gene[qval < 0.10]),
    .groups = "drop"
  ) %>%
  arrange(cell_type, contrast)

head(summary_by_ct_contrast, 12)
```

### Step 5: Visualize top DEGs per cell type

The function `plot_DEG()` selects the top genes per cell type by best
q-value, and draws two side-by-side heatmaps per cell type:

-   the **effect size** (here the `test_statistic`) across all clone
    comparisons, and
-   the matching **-log10(p)** heatmap (capped at `qcap_val`) using the
    same row/column ordering.

`plot_DEG()` returns a named list of the per-cell-type heatmap grobs
(invisibly), so you can either save them to a PDF and/or render them
inline in R.

To save the full set of heatmaps to a PDF, pass a `pdf_path`:

```{r STEP6_plot_deg_pdf, eval=F, fig.width=10, fig.height=5}
source("mixture_DEG_pathway_functions.R")

pdf_path = "example_data/results/celltype_deg_top20_2sided_MTremoved.pdf"
deg_plots <- plot_DEG(all_contrasts,
         cell_types = cell_types,
         pdf_path = pdf_path,
         effect_col = "test_statistic",
         top_k = 20,      # top genes per cell type
         q_thresh = 0.1,  # significance threshold
         qcap_val = 10)   # cap -log10(p) for plotting
```

The full set of per-cell-type heatmaps is saved at
`example_data/results/celltype_deg_top20_2sided_MTremoved.pdf`.

Alternatively, omit `pdf_path` to render the heatmaps directly in R (e.g.
inline in this document). `plot_DEG()` returns the list of grobs, which
you can also redraw individually with `grid::grid.draw()`:

```{r STEP6_plot_deg_inline, eval=F, fig.width=10, fig.height=5}
source("mixture_DEG_pathway_functions.R")

deg_plots <- plot_DEG(all_contrasts,
         pdf_path = pdf_path,
         cell_types = cell_types,
         effect_col = "test_statistic",
         top_k = 20,
         q_thresh = 0.1,
         qcap_val = 10)   # draws each cell type's heatmap inline

```

Or the DEGs specific to a celltype in the clones can be visualized as:
```{r STEP6_plot_deg_in_R, eval=T,fig.width=15, fig.height=8, out.width="80%"}
source("mixture_DEG_pathway_functions.R")

deg_plots <- plot_DEG(all_contrasts, cell_types = cell_types, effect_col = "test_statistic",top_k = 20, q_thresh = 0.1,qcap_val = 10,draw=F)

grid.newpage()
# visualize a specific celltype
grid.draw(deg_plots[["BE"]])
```


### References

[1] Miller, T.E., Lareau, C.A., Verga, J.A. et al. Mitochondrial variant
enrichment from high-throughput single-cell RNA sequencing resolves
clonal populations. *Nat Biotechnol* 40, 1030--1034 (2022).
<https://doi.org/10.1038/s41587-022-01210-8>

[2] Cable, Dylan M., et al. "Robust decomposition of cell type mixtures
in spatial transcriptomics." *Nature Biotechnology* 40.4 (2022):
517-526.

[3] spotGLM. <https://kaishumason.github.io/SpotGLM/>
